---
title: "Adventures in Overfitting!"
author: Michael Schatz (mschatz@jhu.edu)
date: April 20, 2017
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: paper
    keep_md: true
---


# Generate some noisy data
*********************************************************************************************

## Generate the true relationship

```{r}
## Initialize Random Number Generator so plots are consistent
set.seed(20)

## Now compute points along a cubic function
q <- seq(from=-15, to=15, by=.5)
y <- q*q*q-3*q*q-144*q+432
plot(q,y,col='firebrick1',lwd=3, typ="l", ylim=c(min(y)-200, max(y)+200))
```



## Generate random noise

```{r}
noise <- rnorm(length(q), mean=0, sd=333)
hist(noise)
```

## Add noise to original dataset

```{r}
noisy.y <- y + noise
plot(q,y,col='firebrick1',lwd=3, typ="l", ylim=c(min(y)-200, max(y)+200))
points(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)
```


# Fitting Lines to (multiple) points
*********************************************************************************************

##  Given any two points I can draw a straight line between them!

### Start by trying the first and last point

```{r}
i1 <- 1
i2 <- length(q)

q2 <- c(q[i1],q[i2])
ny2 <- c(noisy.y[i1], noisy.y[i2])
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)
points(q2,ny2,col='red',xlab='q',main='Observed data', pch=19)

model1 <- lm(ny2 ~ q2)
abline(a=model1$coefficients[1], b=model1$coefficients[2], col="green", lwd=3)
```



### pick another pair

```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)
points(q2,ny2,col='red',xlab='q',main='Observed data', pch=19)
abline(a=model1$coefficients[1], b=model1$coefficients[2], col="green", lwd=3)

i1 <- 1
i2 <- floor(length(q)/2)

q2 <- c(q[i1],q[i2])
ny2 <- c(noisy.y[i1], noisy.y[i2])

model1 <- lm(ny2 ~ q2)
points(q2,ny2,col='red',xlab='q',main='Observed data', pch=19)
abline(a=model1$coefficients[1], b=model1$coefficients[2], col="green", lwd=3)
```



### find lines to all other points
```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)
lines(q,y,col='firebrick1',lwd=3)

for (i2 in seq(2,length(q))){
  q2 <- c(q[i1],q[i2])
  ny2 <- c(noisy.y[i1], noisy.y[i2])
  model1 <- lm(ny2 ~ q2)

  points(q2,ny2,col='red',xlab='q',main='Observed data', pch=19)
  abline(model1, col="green", lwd=3)
}
```





## Given any three points, I can draw a quadradic between them!

### First, last, middle points

```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)

i1 <- 1
i2 <- floor(length(q)/2)
i3 <- length(q)

q3 <- c(q[i1],q[i2],q[i3])
ny3 <- c(noisy.y[i1], noisy.y[i2], noisy.y[i3])
points(q3,ny3,col='red',pch=19)

model2 <- lm(ny3 ~ poly(q3, 2, raw=TRUE))
model2.y <- model2$coefficients[3]*q*q + model2$coefficients[2]*q + model2$coefficients[1]
lines(q, model2.y, col="green",lwd=3)
```


### Pick 3 new points

```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)

i1 <- 1
i3 <- floor(3*length(q)/4)
i2 <- floor(i3/2)

q3 <- c(q[i1],q[i2],q[i3])
ny3 <- c(noisy.y[i1], noisy.y[i2], noisy.y[i3])
points(q3,ny3,col='red',pch=19)

model2 <- lm(ny3 ~ poly(q3, 2, raw=TRUE))
model2.y <- model2$coefficients[3]*q*q + model2$coefficients[2]*q + model2$coefficients[1]
lines(q, model2.y, col="green",lwd=3)
```



### Fit a bunch of triples

```{r}

plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)

for (i3 in seq(4,length(q))){
  i1 <- 1
  i2 <- floor(i3/2)

  q3 <- c(q[i1],q[i2],q[i3])
  ny3 <- c(noisy.y[i1], noisy.y[i2], noisy.y[i3])

  model2 <- lm(ny3 ~ poly(q3, 2, raw=TRUE))
  model2.y <- model2$coefficients[3]*q*q + model2$coefficients[2]*q + model2$coefficients[1]
  points(q3,ny3,col='red',pch=19)
  lines(q, model2.y, col="green",lwd=3)
}
```



## Given any four points, I can draw a cubic between them!


### First, 1/3, 2/3, last

```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)

i1 <- 1
i2 <- floor(length(q)/3)
i3 <- floor(2*length(q)/4)
i4 <- length(q)

q4 <- c(q[i1],q[i2],q[i3], q[i4])
ny4 <- c(noisy.y[i1], noisy.y[i2], noisy.y[i3], noisy.y[i4])
points(q4,ny4,col='red',pch=19)

model3 <- lm(ny4 ~ poly(q4, 3, raw=TRUE))
model3.y <- model3$coefficients[4]*q*q*q + 
            model3$coefficients[3]*q*q + 
            model3$coefficients[2]*q + 
            model3$coefficients[1]
lines(q, model3.y, col="green",lwd=3)
```

### Shifted First, 1/3, 2/3, last

```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)

i1 <- 10
i2 <- floor(length(q)/3)
i3 <- floor(2*length(q)/4)
i4 <- length(q)-10

q4 <- c(q[i1],q[i2],q[i3], q[i4])
ny4 <- c(noisy.y[i1], noisy.y[i2], noisy.y[i3], noisy.y[i4])
points(q4,ny4,col='red',pch=19)

model3 <- lm(ny4 ~ poly(q4, 3, raw=TRUE))
model3.y <- model3$coefficients[4]*q*q*q + 
            model3$coefficients[3]*q*q + 
            model3$coefficients[2]*q + 
            model3$coefficients[1]
lines(q, model3.y, col="green", lwd=3)
```



## Given n points, I can draw a n-1 degree polynomial between them!


### Pick 10 points at random

```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)

n <- 10
i <- sort(sample(1:length(q), n))

qq <- q[i]
yy <- noisy.y[i]

points(qq,yy,col='red', pch=19)
modeln <- lm(yy ~ poly(qq, n-1, raw=TRUE))
py <- predict(modeln, data.frame(qq=q))
lines(q, py, col="green",lwd=3)
```


### rescale so you can see the full function

```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q', main='Observed data', pch=19,
     ylim=c(min(noisy.y)*30, max(noisy.y)*10))
points(qq,yy,col='red', pch=19)
lines(q, py, col="green",lwd=3)
```





# Linear and Polynomial Regression
*********************************************************************************************

## Lets do something more useful and find the best fit line through the points

### Best linear model
```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)
lines(q,y,col='firebrick1',lwd=3, typ="l", ylim=c(min(y)-200, max(y)+200))
model1 <- lm(noisy.y ~ poly(q,1,raw=TRUE))
yy1    <- predict(model1)
lines(q,yy1, col="green", lwd=3)
```


### Best quadradic model
```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)
lines(q,y,col='firebrick1',lwd=3, typ="l", ylim=c(min(y)-200, max(y)+200))
model2 <- lm(noisy.y ~ poly(q,2,raw=TRUE))
yy2    <- predict(model2)
lines(q,yy2, col="orange", lwd=3)
```

### Best cubic model
```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)
lines(q,y,col='firebrick1',lwd=3, typ="l", ylim=c(min(y)-200, max(y)+200))
model3 <- lm(noisy.y ~ poly(q,3,raw=TRUE))
yy3    <- predict(model3)
lines(q,yy3, col="gold", lwd=3)
```


### Best 30th degree model
```{r}
plot(q,noisy.y,col='deepskyblue4',xlab='q',main='Observed data', pch=19)
lines(q,y,col='firebrick1',lwd=3, typ="l", ylim=c(min(y)-200, max(y)+200))
model30 <- lm(noisy.y ~ poly(q,30,raw=TRUE))
yy30    <- predict(model30)
lines(q,yy30, col="purple", lwd=3)
```

## Final Thoughts

The 30 degree model "matches" the data better than the 3rd degree, but is clearly overfit
and probably wont do a good job representing any new data that we collect. In addition to finding 
curves that smooth datasets, these types of lines are often used to predict disease outcomes
(above curve=cancer, below=benign). If we overfit the boundary to a dataset, we probably
wont do a very good job classifying new cases as they come in. 

*Remember: The chances for overfitting increase with the number of parameters used!*

